---
title: "Studying teacher cognitive load using eyetracking"
output: html_document
bibliography: eyetracking.bibtex
---

This small document contains the main code and results of our analysis of...

We have used recordings from different subjects and kinds of formal education settings:

* Three collaborative learning sessions where primary school students used a tabletop tangible interface to learn fractions. The "teacher" was a researcher with relatively little teaching experience (marked T1S2, T1S3, T1S6)
* Two lectures recorded in (computer science) university courses, given by two different (expert) teachers (marked T2S1, T3S1)
* ...

The following sections explain the process of analysis followed ...

Pre-requisites and data pre-processing
-------------------------------------------

The main input for the following analyses is the raw event data from the recording of the sessions (in the SMI mobile eye-trackers, use the menu ...), as well as saccade and fixation event details, also exported as comma-separated files (use the menu ...).

1. From these raw data files, we extract a sequence of eye-tracking events, which we use as a base to estimate the subjects' cognitive load, namely, the pupil diameter, fixation length and saccade speed
2. Then, from these events, we extract the four metrics that, based on @Buettner2013, can be used to estimate the cognitive load, over sliding windows of 10s, with 5 second step/slide:
    * Average pupil diameter
    * Standard deviation of the pupil diameter
    * Number of fixations >500ms
    * Average saccade speed

```{r, cache=TRUE, echo=TRUE, message=FALSE}
# Some basic parameters for the sliding windows (in seconds)
window <- 10
slide <- 5

# (in Mirko's original labeling, Pierre02 corresponds to T2S1, Roland03 is T3S1)
sessions <- c("T2S1","T3S1")
for (session in sessions){
    
    # We check whether the clean data is already in place - if so, we skip this pre-processing
    if(!file.exists(paste("./rawdata/",session,".EyetrackerEvents.rda",sep="")) ||
        !file.exists(paste("./rawdata/",session,".EyetrackerFixations.rda",sep="")) ||
        !file.exists(paste("./rawdata/",session,".EyetrackerSaccades.rda",sep=""))){
        
        # We load the raw events export
        filename = paste("./rawdata/",session,"-events-export.txt", sep="")
        filedata <- read.csv(filename,as.is=T,comment.char="#")
        
        # From all the data, we only need timestamp, pupil diameter (L,R, in mm)
        filedata <- filedata[c(1,6,9)]
        filedata$Session <- session
        pupildata <- data.frame(filedata)
    
        # We calculate the time baseline of the session
        time0 <- min(pupildata$Time)
        pupildata$Time.ms <- (pupildata$Time - time0) / 1000
        
        # We load the fixation details file
        filename = paste("./rawdata/",session,"-fixation-details.txt", sep="")
        filedata <- read.csv(filename,comment.char="#")
        
        # we select the meaningful columns (for now, only fixation start, duration, end in ms)
        filedata <- filedata[c(7,8,9)] # TODO: In the JdC files, it is 8,9,10!!
        filedata$Session <- session
        fixdata <- data.frame(filedata)
    
        #We set the time of the fixation in the middle of the fixation
        fixdata$Time.ms <- (fixdata$Fixation.Start..ms. + (fixdata$Fixation.Duration..ms./2)) 
        # We create a Time field so that we have the time in both timestamp and ms formats
        fixdata$Time <- time0 + (fixdata$Time.ms)*1000
    
        # We load the saccade details file
        filename = paste("./rawdata/",session,"-saccade-details.txt", sep="")
        filedata <- read.csv(filename,comment.char="#")
        
        # we select the meaningful columns (for now, only saccade start, duration, end in ms and amplitude in degrees)
        filedata <- filedata[c(7,8,9,14)] # TODO: In the JdC files, it is 8,9,10,15!!
        filedata$Session <- session
        sacdata <- data.frame(filedata)
        # We add the saccade speed for each saccade
        sacdata$Saccade.Speed <- sacdata$Amplitude.... / sacdata$Saccade.Duration..ms.
    
        #We set the time of saccade in the middle of the fixation
        sacdata$Time.ms <- (sacdata$Saccade.Start..ms. + (sacdata$Saccade.Duration..ms./2)) 
        # We create a Time field so that we have the time in both timestamp and ms formats
        sacdata$Time <- time0 + (sacdata$Time.ms)*1000
        
        # We save the clean(er) data to smaller, more efficient rda files
        save(pupildata,file=paste("./rawdata/",session,".EyetrackerEvents.rda",sep=""),compress=TRUE)
        save(fixdata,file=paste("./rawdata/",session,".EyetrackerFixations.rda",sep=""),compress=TRUE)
        save(sacdata,file=paste("./rawdata/",session,".EyetrackerSaccades.rda",sep=""),compress=TRUE)
        
    }
    
    # We load the clean data, just in case we did not the previous steps
    pupildata <- get(load(paste("./rawdata/",session,".EyetrackerEvents.rda",sep="")))
    fixdata <- get(load(paste("./rawdata/",session,".EyetrackerFixations.rda",sep="")))
    sacdata <- get(load(paste("./rawdata/",session,".EyetrackerSaccades.rda",sep="")))

    # We load a few functions that will be useful for the rolling window calculations
    source("JDCEyetrackingAnalysis.R")
    
    # We get the rolling window for the mean pupil diameter, and its median value for a median cut
    meandata <- rollingMean(pupildata$Time.ms,pupildata$L.Pupil.Diameter..mm.,window*1000,slide*1000)
    meanPDmedian <- median(meandata$value)

    # We get the rolling window for the SD of pupil diameter, and its median value for a median cut
    sddata <- rollingSd(pupildata$Time.ms,pupildata$L.Pupil.Diameter..mm.,window*1000,slide*1000)
    sdPDmedian <- median(sddata$value)

    # We get the number of long fixations in the window, and its median
    longdata <- rollingLong(fixdata$Time.ms,fixdata$Fixation.Duration..ms.,window*1000,slide*1000)
    longFixMedian <- median(longdata$value)

    # We get the saccade speed in the window
    sacspdata <- rollingMean(sacdata$Time.ms,sacdata$Saccade.Speed,window*1000,slide*1000)
    sacSpMedian <- median(sacspdata$value)
    
    # We save all the (clean) window data in a file, for later use
    save(meandata,meanPDmedian,sddata,sdPDmedian,longdata,longFixMedian,sacspdata,sacSpMedian,file=paste("./cleandata/",session,".CognitiveLoadMetrics.rda",sep=""))
    
}
```

After doing this pre-processing, we have four data frames per each session, looking like:

```{r}
session <- sessions[1]
load(paste("./cleandata/",session,".CognitiveLoadMetrics.rda",sep=""))
head(meandata, n=3)
head(sddata, n=3)
head(longdata, n=3)
head(sacspdata, n=3)
```

Analysis for T2S1 (Pierre)
------------------------------------------------




References
--------------------------------
